---
title: "Romance-Novel-Data-Visualization-Workshop"
output: pdf_document
date: "2026-01-22"
---

**Learning Objectives**:
-- 

**What Is The Goal of Data Visualization?** 
* --Visual Proof--: A great data visualization provides proof that our data shows what we have described in words. You can certainly take what I say at face value, but critically, you should expect me to show it to you. As scientist, we strive to be worthy of our readers' trust and data visualizations can assist with that.
* --Data Storytelling--: A picture is worth 1000 words! If you generate a great visualization it can adequately communicate many pieces of information. Let's say we are looking at trends in sales of different colored widgets: red widgets, blue widgets, green widgets, and yellow widgets. Run the code chunk below to generate our dummy data set of widget information: 

```{r}
## Load a special R package that will help us format the data frame
library(tidyverse)

## set a seed so the data does not change between my run of this code
set.seed(1234)

## Set our time period 2020-2025
years <- 2020:2025

widget_data <- expand_grid(
  year = years,
  color = c("Yellow", "Blue", "Red", "Green")
) %>%
  mutate(
    ## Setting sales with different trends by color
    sales = case_when(
      color == "Yellow" ~ 500 - (year - 2020) * 30 + rnorm(n(), 0, 20),
      color == "Blue" ~ 200 + (year - 2020) * 40 + rnorm(n(), 0, 15),
      color == "Red" ~ 150 + (year - 2020) * 50 + rnorm(n(), 0, 15),
      color == "Green" ~ 100 + (year - 2020) * 60 + rnorm(n(), 0, 15)
    ),
    sales = pmax(sales, 0), ## Ensure no negative sales
    sales = round(sales)
  )

## View the data
head(widget_data)
```
Now that we have generated our dummy data, let's generate a visualization that explores whether or not sales of widgets increased or decreased over the 2020-2025 perid. We could plot a single line showing how many people purchased widgets between 2020 and 2025. Take a look at the visualization shown below. What does it tell us about our data? 


```{r}
# Aggregate total sales per year
total_sales <- widget_data %>%
  group_by(year) %>%
  summarise(sales = sum(sales))

plot1 <- ggplot(total_sales, aes(x = year, y = sales)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  labs(
    title = "Total Widget Sales (2020-2025)",
    subtitle = "All Widgets",
    x = "Year",
    y = "Number of Sales"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(plot1)
```

More appropriately, I could plot four different lines, each for the sale of different colored widgets. This plot shows me that the  number of sales for my red, blue, and green widgets increases over the 2020-2025 time period, but the sales for my yellow widgets decreases! We wouldn't have known that had we not broken down or visualization by the groups within our "Color" variable. By the way, this reversal of trends when looking at subgroups of our variable is known as **Simpson's Paradox**. 

```{r}
plot2 <- ggplot(widget_data, aes(x = year, y = sales, color = color, group = color)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Widget Sales by Color (2020-2025)",
    subtitle = "Individual Trends by Widget Color",
    x = "Year",
    y = "Number of Sales",
    color = "Widget Color"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold")) +
  scale_color_manual(values = c("Yellow" = "#FFD700", "Blue" = "#4169E1", 
                                 "Red" = "#DC143C", "Green" = "#228B22"))

print(plot2)
```
* Presentation Purposes: No one wants to read all 2000 words on your point point slide during a 15 minute presentation of your results. A visualization will not only summarize your information, but will clearly communicate your findings in a memorable way. Humans remember things that they see! Some of my earliest memories are not auditory, but rather visual. I remember the color of shirt my Grandma wore when she came to pick me up from shool one day when I was a kid. 

What makes a visualization good? 
* Clear titles and labels
* Clearly distinguishable colors - use color blind friendly palettes
* Tells a data story without being a cluttered appearance
* Think about your audience! Would you show your 5 year old cousin the same graphic you would show your work collegue? 

How can we create good visualizations?
* Think about the question that you are trying to solve. 
* Think about the answer you found in your data. 
* How can you best tell the story of what you found using a picture?

What tools are at our disposal to generate visualizations in R?
* Base R
* ggplot2

Kaggle - Romance Novel Data Set
* Emphasize how wonderful data science is that you have the opportunity to explore any field at all. Finance, Literature, Physics, Biology, Anthropology, etc. You can apply data science, especially data visualization to express
* Column Names
* Data Dimensions
* Data Types
* Assessment of Missingness

Base R Visualizations - Let's Eyeball It
* Histograms
* 
```{r setup, include=TRUE}

```
